Stellar Radiation is Critical for Regulating Star Formation and Driving Outflows in Low Mass Dwarf Galaxies

### 2019-04-08 thesis/chapters/ch_chapter2.tex
Fixing figure size and labels

\begin{figure*}
\centering
\includegraphics[width=0.49\linewidth]{figures/ch2/sfr}
\includegraphics[width=0.49\linewidth]{figures/ch2/mass}
\caption{The star formation rate (left) and gas and stellar mass (right) of each of our three simulations over time. Time bins with no star formation are left empty.}
\label{ch2:figsfr_mass_evolution}
\end{figure*}

We compare the resulting star formation rate (left) and gas mass properties (right) of our three simulations in Figure~\ref{ch2:figsfr_mass_evolution}. There is a clear, significant contrast between the runs with and without ionizing radiation. Stellar ionizing radiation leads to a factor of $\sim$ 5 reduction in the resulting SFR, as compared to the noRT run. Since the shortrad and fiducial simulations are so similar over the first $\sim$~100~Myr, it is clear that stellar radiation acts to significantly reduce the local star formation efficiency around young, massive stars. Radiation from these stars quickly ionizes and dissipates surrounding dense gas that would otherwise have gone to fuel a significant amount of additional star formation. This is confirmed in Figure~\ref{ch2:figsf gas}, which shows the gas mass in each simulation above the SF density threshold of $n = 200$~cm$^{-3}$ during the first 100~Myr. While our fiducial and shortrad simulations remain roughly the same here, the noRT run, at its peak, has an order of magnitude more cold, dense gas. \footnote{We are unable to follow the long-term evolution of the noRT galaxy due to computational constraints. Although radiative transfer itself is computationally expensive, this run is substantially more costly due to a lower typical timestep and increased cost in computing the optically thin radiation effects (photoelectric effect and LW dissociation) for the additional star particles.}

\begin{figure}
\centering
\includegraphics[width=0.65\linewidth]{figures/ch2/mass_density_cut}
\caption{The total gas mass at $n > 200$~cm$^{-3}$, our star formation density threshold.}
\label{ch2:figsf gas}
\end{figure}

@@

This can be confirmed by examining the metal retention fraction (the fraction of produced metals retained within the disk of the galaxy) and mass outflow rates across simulations. As shown in Figure~\ref{ch2:figmetal_retention}, the mass outflow rate in the fiducial run peaks at an order of magnitude higher at 0.25~R$_{\rm vir}$ than the shortrad simulation, declining only due to a comparative drop off in star formation. The outflow in noRT is only a factor of a few lower than the fiducial run, but it requires a five times higher supernova rate in this simulation to match the same outflow seen with full stellar radiation feedback, so it implies a far lower mass loading factor. Because of this difference in SFR,
%mm [rewrote to display equation]
the differences across simulations are more significant for our computed mass loading factor
\begin{equation}
    \eta = \frac{ \dot{M}_{\rm out}}{ <\rm{SFR}>_{100 \rm{Myr}}}.
\end{equation} The brackets indicate time averaging over 100~Myr.
%
While the fiducial run reaches a peak $\eta$ of a few hundred, shortrad is consistently an order of magnitude or more below this. The noRT simulation is even lower.

@@

\begin{figure*}
\centering
\includegraphics[width=0.32\linewidth]{figures/ch2/metal_retention}
\includegraphics[width=0.32\linewidth]{figures/ch2/mass_outflow_comparison}
\includegraphics[width=0.32\linewidth]{figures/ch2/mass_loading_comparison}
\caption{{\bf Left:} The fraction of metals contained within the ISM of each galaxy (blue) and the fraction ejected beyond the virial radius (orange). As in Figure~\ref{ch2:figsfr_mass_evolution}, the fiducial run is given as solid lines, while the noRT and shortrad are dash-dotted and dashed respectively. {\bf Middle:} The mass outflow rate for each run at 0.25~R$_{\rm vir}$ (solid) and R$_{\rm vir}$ (dashed). {\bf Right:} The mass loading factor $\eta$ calculated as the outflow rate divided by the SFR averaged over 100~Myr. Inclusion of diffuse ionization makes an order of magnitude difference in $\eta$ in this case. }
\label{ch2:figmetal_retention}
\end{figure*}

\section{Discussion} \label{ch2:sec:discussion}
Accounting for feedback from stellar radiation plays a significant role in determining the ability for SN energy to couple to the ISM and therefore drive outflows. We believe this work is novel in examining the
importance of localized ionization vs.\ ionization from a diffuse radiation field far from a single star. Modeling only local stellar ionizing feedback is insufficient to describe the long-term evolution of an isolated dwarf galaxy. To explore the cause of this difference we compare in Figure~\ref{ch2:figpanel1} and~Figure~\ref{ch2:figpanel2} the gas number density (left), temperature (middle), and hydrogen ionization fraction (right) in edge-on slices in each of our simulations at two different times.

\begin{figure*}
\centering
\includegraphics[width=0.99\linewidth]{figures/ch2/DD0136_fiducial_shortrad_nort}
\caption{Edge-on slices of each simulation showing gas number density (left), temperature (middle), and hydrogen ionization fraction (right) 17~Myr after the formation of the first star in each run. Each panel is 4 kpc x 4 kpc.
%The full movie of this evolution from 0 to 118 Myr is included for further clarification.
See \cite{Emerick2018a} for a movie of this evolution.}
%The movie demonstrates that the differences in outflows between the simulations depends on the ability for feedback to carve low-density, ionized channels along which SNe can launch gas beyond the disk of the galaxy. These channels are readily created by ionizing radiation in the fiducial simulation, or to some degree by SNe in the noRT simulation due to its much higher SN rate. However, SNe launched gas is trapped by neutral, cold shells in the shortrad simulation, unable to easily escape from the disk of the simulated galaxy. These same insights can also be gained by comparing this figure with Figure~\ref{ch2:figpanel2}.}
\label{ch2:figpanel1}
\end{figure*}

\begin{figure*}
\centering
\includegraphics[width=0.99\linewidth]{figures/ch2/DD0160_fiducial_shortrad_nort}
\caption{Same as Figure~\ref{ch2:figpanel1}, but at 40~Myr.}
\label{ch2:figpanel2}
\end{figure*}

Figure~\ref{ch2:figpanel1}, at 17~Myr, compares each simulation just after the first few SNe. Already there are significant differences between the runs. Gas outside the galaxy is warm ($\sim$~10$^{4}$~K) and ionized up to $\sim$500~pc above/below the plane of the disk in our fiducial run. This same gas is cold ($<10^4$~K) and neutral in both other runs. The contrast between the effect of ionizing radiation in the fiducial and shortrad runs is seen most clearly in the ionized region in-plane and to the right of center. This region contains massive stars that are capable of generating enough ionizing radiation to carve a channel out to the halo of our fiducial simulation; this does not occur in the shortrad case. Instead, the HII region is confined by surrounding cold, neutral gas.

Although the ISM properties within the HII region in each case are quite similar between the simulations, the SNe that eventually go off in this region are confined by the neutral gas in the shortrad simulation, but readily escape through the lower density ionized gas into the galaxy halo in our fiducial case. As these simulations evolve, the existence of these diffuse, ionized channels in the fiducial run easily allow continual and significant outflows from SNe, as shown in  Figure~\ref{ch2:figpanel2}, which shows each simulation 40~Myr after the formation of the first stars. In contrast, the same SNe in the other two simulations are well contained, surrounded by shells of denser, neutral gas. Though they make an important local impact on the ISM, they are unable to drive significant mass loss from the galaxy. In the noRT case, an outflow does eventually develop, but it takes a factor of five increase in SFR, and a corresponding increase in supernova rate, for SNe to finally break out from the neutral gas surrounding the galaxy.

As shown in Figure~\ref{ch2:figmetal_retention} the differences in ionization structure and its effect on galactic winds has direct consequences for the chemical evolution of our dwarf galaxy. The winds in our fiducial simulation carry nearly all of the metals produced out from the disk of our galaxy. This is the only run in agreement with observations of Local Group dwarf galaxies, both dSph's \citep{Kirby2011} and the gaseous, star forming Local Group dwarf galaxy Leo P \citep{McQuinn2015}, with metal ejection fractions of up to 95\%. This is in stark contrast to the $\sim$30\% retention fraction in our shortrad simulation. This would also influence the chemical enrichment of neighboring galaxies, given the significant differences in metal ejection past the virial radius between these two runs. Clearly the effects of feedback on observable chemical properties of galaxies is a key discriminator among models.

>>>>

\begin{figure*}
\centering
\includegraphics[width=0.49\linewidth]{figures/ch2/sfr}
\includegraphics[width=0.49\linewidth]{figures/ch2/mass}
\caption{The star formation rate (left) and gas and stellar mass (right) of each of our three simulations over time. Time bins with no star formation are left empty.}
\label{ch2:fig:sfr_mass_evolution}
\end{figure*}

We compare the resulting star formation rate (left) and gas mass properties (right) of our three simulations in Figure~\ref{ch2:fig:sfr_mass_evolution}. There is a clear, significant contrast between the runs with and without ionizing radiation. Stellar ionizing radiation leads to a factor of $\sim$ 5 reduction in the resulting SFR, as compared to the noRT run. Since the shortrad and fiducial simulations are so similar over the first $\sim$~100~Myr, it is clear that stellar radiation acts to significantly reduce the local star formation efficiency around young, massive stars. Radiation from these stars quickly ionizes and dissipates surrounding dense gas that would otherwise have gone to fuel a significant amount of additional star formation. This is confirmed in Figure~\ref{ch2:fig:sf gas}, which shows the gas mass in each simulation above the SF density threshold of $n = 200$~cm$^{-3}$ during the first 100~Myr. While our fiducial and shortrad simulations remain roughly the same here, the noRT run, at its peak, has an order of magnitude more cold, dense gas. \footnote{We are unable to follow the long-term evolution of the noRT galaxy due to computational constraints. Although radiative transfer itself is computationally expensive, this run is substantially more costly due to a lower typical timestep and increased cost in computing the optically thin radiation effects (photoelectric effect and LW dissociation) for the additional star particles.}

\begin{figure}
\centering
\includegraphics[width=0.65\linewidth]{figures/ch2/mass_density_cut}
\caption{The total gas mass at $n > 200$~cm$^{-3}$, our star formation density threshold.}
\label{ch2:fig:sf gas}
\end{figure}

@@

This can be confirmed by examining the metal retention fraction (the fraction of produced metals retained within the disk of the galaxy) and mass outflow rates across simulations. As shown in Figure~\ref{ch2:fig:metal_retention}, the mass outflow rate in the fiducial run peaks at an order of magnitude higher at 0.25~R$_{\rm vir}$ than the shortrad simulation, declining only due to a comparative drop off in star formation. The outflow in noRT is only a factor of a few lower than the fiducial run, but it requires a five times higher supernova rate in this simulation to match the same outflow seen with full stellar radiation feedback, so it implies a far lower mass loading factor. Because of this difference in SFR,
%mm [rewrote to display equation]
the differences across simulations are more significant for our computed mass loading factor
\begin{equation}
    \eta = \frac{ \dot{M}_{\rm out}}{ <\rm{SFR}>_{100 \rm{Myr}}}.
\end{equation} The brackets indicate time averaging over 100~Myr.
%
While the fiducial run reaches a peak $\eta$ of a few hundred, shortrad is consistently an order of magnitude or more below this. The noRT simulation is even lower.

@@

\begin{figure*}
\centering
\includegraphics[width=0.45\linewidth]{figures/ch2/metal_retention}
\includegraphics[width=0.45\linewidth]{figures/ch2/mass_outflow_comparison}\\
\includegraphics[width=0.45\linewidth]{figures/ch2/mass_loading_comparison}
\caption{{\bf Top Left:} The fraction of metals contained within the ISM of each galaxy (blue) and the fraction ejected beyond the virial radius (orange). As in Figure~\ref{ch2:fig:sfr_mass_evolution}, the fiducial run is given as solid lines, while the noRT and shortrad are dash-dotted and dashed respectively. {\bf Top Right:} The mass outflow rate for each run at 0.25~R$_{\rm vir}$ (solid) and R$_{\rm vir}$ (dashed). {\bf Bottom:} The mass loading factor $\eta$ calculated as the outflow rate divided by the SFR averaged over 100~Myr. Inclusion of diffuse ionization makes an order of magnitude difference in $\eta$ in this case. }
\label{ch2:fig:metal_retention}
\end{figure*}

\section{Discussion} \label{ch2:sec:discussion}
Accounting for feedback from stellar radiation plays a significant role in determining the ability for SN energy to couple to the ISM and therefore drive outflows. We believe this work is novel in examining the
importance of localized ionization vs.\ ionization from a diffuse radiation field far from a single star. Modeling only local stellar ionizing feedback is insufficient to describe the long-term evolution of an isolated dwarf galaxy. To explore the cause of this difference we compare in Figure~\ref{ch2:fig:panel1} and~Figure~\ref{ch2:fig:panel2} the gas number density (left), temperature (middle), and hydrogen ionization fraction (right) in edge-on slices in each of our simulations at two different times.

\begin{figure*}
\centering
\includegraphics[width=0.99\linewidth]{figures/ch2/DD0136_fiducial_shortrad_nort}
\caption{Edge-on slices of each simulation showing gas number density (left), temperature (middle), and hydrogen ionization fraction (right) 17~Myr after the formation of the first star in each run. Each panel is 4 kpc x 4 kpc.
%The full movie of this evolution from 0 to 118 Myr is included for further clarification.
See \cite{Emerick2018a} for a movie of this evolution.}
%The movie demonstrates that the differences in outflows between the simulations depends on the ability for feedback to carve low-density, ionized channels along which SNe can launch gas beyond the disk of the galaxy. These channels are readily created by ionizing radiation in the fiducial simulation, or to some degree by SNe in the noRT simulation due to its much higher SN rate. However, SNe launched gas is trapped by neutral, cold shells in the shortrad simulation, unable to easily escape from the disk of the simulated galaxy. These same insights can also be gained by comparing this figure with Figure~\ref{ch2:fig:panel2}.}
\label{ch2:fig:panel1}
\end{figure*}

\begin{figure*}
\centering
\includegraphics[width=0.99\linewidth]{figures/ch2/DD0160_fiducial_shortrad_nort}
\caption{Same as Figure~\ref{ch2:fig:panel1}, but at 40~Myr.}
\label{ch2:fig:panel2}
\end{figure*}

Figure~\ref{ch2:fig:panel1}, at 17~Myr, compares each simulation just after the first few SNe. Already there are significant differences between the runs. Gas outside the galaxy is warm ($\sim$~10$^{4}$~K) and ionized up to $\sim$500~pc above/below the plane of the disk in our fiducial run. This same gas is cold ($<10^4$~K) and neutral in both other runs. The contrast between the effect of ionizing radiation in the fiducial and shortrad runs is seen most clearly in the ionized region in-plane and to the right of center. This region contains massive stars that are capable of generating enough ionizing radiation to carve a channel out to the halo of our fiducial simulation; this does not occur in the shortrad case. Instead, the HII region is confined by surrounding cold, neutral gas.

Although the ISM properties within the HII region in each case are quite similar between the simulations, the SNe that eventually go off in this region are confined by the neutral gas in the shortrad simulation, but readily escape through the lower density ionized gas into the galaxy halo in our fiducial case. As these simulations evolve, the existence of these diffuse, ionized channels in the fiducial run easily allow continual and significant outflows from SNe, as shown in  Figure~\ref{ch2:fig:panel2}, which shows each simulation 40~Myr after the formation of the first stars. In contrast, the same SNe in the other two simulations are well contained, surrounded by shells of denser, neutral gas. Though they make an important local impact on the ISM, they are unable to drive significant mass loss from the galaxy. In the noRT case, an outflow does eventually develop, but it takes a factor of five increase in SFR, and a corresponding increase in supernova rate, for SNe to finally break out from the neutral gas surrounding the galaxy.

As shown in Figure~\ref{ch2:fig:metal_retention} the differences in ionization structure and its effect on galactic winds has direct consequences for the chemical evolution of our dwarf galaxy. The winds in our fiducial simulation carry nearly all of the metals produced out from the disk of our galaxy. This is the only run in agreement with observations of Local Group dwarf galaxies, both dSph's \citep{Kirby2011} and the gaseous, star forming Local Group dwarf galaxy Leo P \citep{McQuinn2015}, with metal ejection fractions of up to 95\%. This is in stark contrast to the $\sim$30\% retention fraction in our shortrad simulation. This would also influence the chemical enrichment of neighboring galaxies, given the significant differences in metal ejection past the virial radius between these two runs. Clearly the effects of feedback on observable chemical properties of galaxies is a key discriminator among models.
